\large{\bf Looking forwards and backwards through locally regulated populations }

### 2022-05-02 forwards-and-backwards.tex
Update on Overleaf.

@@

Although our proof of Theorem~\ref{thm:lineages}
uses the explicit representation in terms of the lookdown process,
informal calculations agree with the main result.
Suppose that we are tracking a lineage that is at location $y$ at time $t$.
Looking back through time, the lineage will move at the time that individual was born,
and will jump to the location of the parent of that individual.
Now, the rate at which new individuals are born to parents at $x$ and establish at $y$
is
$$
    \gamma(x, \eta_t) q(x, dy) r(y, \eta_t) .
$$
Happily, we are interested in the case when $\eta$ has a density,
say $\eta_t(dx) = \varphi_t(x) dx$,
so interpreting these rates is not too difficult.
This implies that the probability a randomly chosen individual near $y$
is a new offspring from a parent at $x$ in $[t, t+dt)$ is, informally,
\begin{equation} \label{eqn:informal_rates}
\frac{
    \gamma(x, \eta_t) q_\theta(x, dy) r(y, \eta_t) \varphi_t(x)
}{
    \varphi_t(y)
} \frac{ q_\theta(x, dy) }{ dy } dx dt .
\end{equation}
Leaving aside questions of whether a lineage can be treated as a randomly chosen individual,
we might then proceed by defining a continuous-time jump process
whose transition rates are given by \eqref{eqn:informal_rates} given $(\varphi_t)_{t=0}^T$.
Let $(L^\theta_s)_{s=0}^T$ be the location of a lineage that moves according to
jump rates~\eqref{eqn:informal_rates},
back through time (i.e., with the substitution that $s = T-t$).
Then, writing $q_\theta(x, y)$ for the density of $q_\theta$,
\begin{align} \label{eqn:lineage_generator_Edt}
    \begin{split}
    &\IE[f(L^\theta_{s+ds}) - f(y) \;|\; L_s = y]
    \\&\qquad 
    =
    ds \int \left(f(x) - f(y)\right)
    \frac{
        \varphi_{T-s}(x) \gamma(x, \eta_{T-s}) r(y, \eta_{T-s})
    }{
        \varphi_{T-s}(y)
    }
    q_\theta(x, y) dx .
    \end{split}
\end{align}
\comment{I'm not sure if at this point in the paper we want to either
(a) write out the next statement; (b) restrict to the $\DG = \Delta$ case,
or (c) leave it as is (as an exercise for the reader, I guess?}
Referring back to Definition~\ref{def:dispersal_generator},
it is not too hard to show that as $\theta \to \infty$,
\begin{align*}
    \theta \int (f(x) - f(y)) q_\theta(x, y) dx
    &\to
    \DG^* f(x)  - f(x) \DG^* 1 .
\end{align*}
Consequently,
\begin{align*}
    &
    \theta \int (f(x) - f(y)) g(x) q_\theta(x, y) dx \\
    &\qquad =
    \theta \int \left\{
        (f(x) g(x) - f(y) g(y)) - f(y) (g(x) - g(y))
    \right\} q_\theta(x, y) dx \\
    &\qquad \to
        \DG^*(fg)(x) - f(x) \DG^* g(x) . 
\end{align*}
Applying this to~\eqref{eqn:lineage_generator_Edt} with $g = \varphi_{T-s} \gamma$,
this suggests that the generator of the limiting process is
\begin{align} \label{eqn:heuristic_lineage_generator}
    \Lgen_s f
    &=
    \frac{r}{\varphi_{T-s}}
    \left\{
        \DG^*(\gamma \varphi_{T-s} f) - f \DG^*(\gamma \varphi_{T-s})
    \right\} .
\end{align}
This agrees with Theorem~\ref{thm:lineages}.

@@

The gneral set-up is as follows.
Each individual will be labeled with a ``level'',
which will be a number in $[0, \lambda]$.
When taking the limit, we will set $\lambda = N$.
\comment{Should we just use $N$ instead of $\lambda$? Or vice-versa?}
We will still encode the process embellished by these levels
as a point measure:
if the $i^\mathrm{th}$ individual's spatial location is $x_i$
and level is $u_i$, then we will write:
$$
    \lp = \sum_i \delta_{x_i, u_i} ,
$$
which is a measure on $\IR^d \times [0, \lambda]$.
Note that the each individual contributes mass 1 to the measure,
not $1/N$ as above.
In some sense, these levels do not affect the process:
the pushforward of this measure-valued ``spatial-level process'',
divided by $N$,
will agree with the purely spatial process defined above.
Furthermore, the levels of the individuals present at any spatial location
conditionally uniform on $[0, \lambda]$.
However, an individual's level in some sense encodes their future reproductive output:
individuals with lower levels tend to live longer, and have more offspring.
For more explanation of the set-up and how this is possible,
see \citet{etheridge/kurtz:2018} and \citet{kurtz/rodrigues:2011}.
Furthermore, the limiting spatial-level process will still be a point measure,
and so we explicitly retain the notion of individuals and lineages through to the 
large-population-density limit.

@@

We can now express our population processes using these levels:
\begin{align*}
    \eta_t = \sum_{a : \tau_a > t \;\&\; U_a(t) < \lambda} \delta_{X_a(t)}
    \qquad \text{and} \qquad
    \lp = \sum_{a : \tau_a > t \;\&\; U_a(t) < \lambda} \delta_{(X_a(t), U_a(t))} .
\end{align*}

@@

Now, let $\beta(x, \eta) = \meanq + 2 \Sigma \grad \log r(x, \eta)$,
and associate with each label an independent $d$-dimensional Brownian motion $W_a$,
and let $\covq = K K^T$.
Then, we can summarize the results above by saying that
in the infinite density limit,
the spatial locations and level of each line of descent $a$
behave for $\tau_a \le t < \sigma_a$ as
\begin{align*}
X_a(t)
    &=
    X_a(\tau_a)
    + \int_{\tau_a}^{t}
    \bigg(
        \beta(x, \eta(s)) ds
        +
        K dW_a(s)
    \bigg)
    ds  \\
U_a(t)
    &=
    U_a(\tau_a)
    + \int_{\tau_a}^{t}
    \bigg(
        \alpha \gamma(X_a(s),\eta(s))
        r(x,\eta(s)) U_a(s)^2
\\ &\qquad \qquad \qquad {}   
        -
        \left\{
            \gamma(X_a(s),\eta(s)) \DG r(X_a(s),\eta(s))
            + F(X_a(s), \eta(s))
        \right\}
        U_a(s)
    \bigg)
    ds .
\end{align*}
To complete the description,
for each label $a \in \labelspace$ let $R_a$ denote an independent Poisson process on $[0, \infty)^2$
with Lebesgue mean measure,
and as before order the points of each in some way.
If the population state at time 0 is $\eta(0)$,
then let $(x_i, u_i)$ be the points of a Poisson process
with mean measure on $\IR^d \times [0, \infty)$ a product of $\eta(0)$ and Lebesgue measure.
\comment{Here I've switched $t$ to be the absolute birth time
rather than relative to the birth time of the parent; I think it's clearer?
It, should be made to agree with the previous section.}
For each point begin a line of descent
with label $i$, location $X_i(0) = x_i$, level $U_a(0) = u_i$, and birth time $\tau_i = 0$.
Each point in the $R_a$ denotes a potential birth time for $a$:
if the $j^\text{th}$ point in $R_a$ is $(t, v)$,
$\tau_a \ge t$, and and $U_a(t) < \infty$,
then a new line of descent with label $a \concat j$ is produced,
with birth time $\tau_{a \concat j} = t$, location $X_{a \concat j}(t) = X_a(t)$, and level
$$
    U_{a \concat j}(t) = U_a(t)
    + \frac{v}{ 2 \alpha \gamma(X_a(t), \eta(t)) r(X_a(t), \eta(t)) } .
$$
\comment{Perhaps the conclusion of the Propostion would be something like:}
We can now express our population processes in the limit using these levels:
\begin{align*}
    \eta_t = \sum_{a : \tau_a > t \;\&\; U_a(t) < \lambda} \delta_{X_a(t)}
    \qquad \text{and} \qquad
    \lp = \sum_{a : \tau_a > t \;\&\; U_a(t) < \lambda} \delta_{(X_a(t), U_a(t))} .
\end{align*}

>>>>

\comment{TODO: talk about $N \to \infty$ rather than $\theta \to \infty$.}

@@

Although our proof of Theorem~\ref{thm:lineages}
uses the explicit representation in terms of the lookdown process,
informal calculations agree with the main result.
Suppose that we are tracking a lineage that is at location $y$ at time $t$.
Looking back through time, the lineage will move at the time that individual was born,
and will jump to the location of the parent of that individual.
Now, the rate at which new individuals are born to parents at $x$ and establish at $y$
is
$$
    \gamma(x, \eta_t) q(x, dy) r(y, \eta_t) .
$$
\comment{TODO: explain this more.}
Happily, we are interested in the case when $\eta$ has a density,
say $\eta_t(dx) = \varphi_t(x) dx$,
so interpreting these rates is not too difficult.
This implies that the probability a randomly chosen individual near $y$
is a new offspring from a parent at $x$ in $[t, t+dt)$ is, informally,
\begin{equation} \label{eqn:informal_rates}
\frac{
    \gamma(x, \eta_t) r(y, \eta_t) \varphi_t(x)
}{
    \varphi_t(y)
} \frac{ q_\theta(x, dy) }{ dy } dx dt .
\end{equation}
Leaving aside questions of whether a lineage can be treated as a randomly chosen individual,
we might then proceed by defining a continuous-time jump process
whose transition rates are given by \eqref{eqn:informal_rates} given $(\varphi_t)_{t=0}^T$.
Let $(L^\theta_s)_{s=0}^T$ be the location of a lineage that moves according to
jump rates~\eqref{eqn:informal_rates},
back through time (i.e., with the substitution that $s = T-t$).
Then, writing $q_\theta(x, y)$ for the density of $q_\theta$,
\begin{align} \label{eqn:lineage_generator_Edt}
    \begin{split}
    &\IE[f(L^\theta_{s+ds}) - f(y) \;|\; L_s = y]
    \\&\qquad 
    =
    ds \int \left(f(x) - f(y)\right)
    \frac{
        \varphi_{T-s}(x) \gamma(x, \eta_{T-s}) r(y, \eta_{T-s})
    }{
        \varphi_{T-s}(y)
    }
    q_\theta(x, y) dx .
    \end{split}
\end{align}
\comment{I'm not sure if at this point in the paper we want to either
(a) write out the next statement; (b) restrict to the $\DG = \Delta$ case,
or (c) leave it as is (as an exercise for the reader, I guess?}
Referring back to Definition~\ref{def:dispersal_generator},
it is not too hard to show that as $\theta \to \infty$,
\begin{align*}
    \theta \int (f(x) - f(y)) q_\theta(x, y) dx
    &\to
    \DG^* f(x)  - f(x) \DG^* 1 .
\end{align*}
Consequently,
\begin{align*}
    &
    \theta \int (f(x) - f(y)) g(x) q_\theta(x, y) dx \\
    &\qquad =
    \theta \int \left\{
        (f(x) g(x) - f(y) g(y)) - f(y) (g(x) - g(y))
    \right\} q_\theta(x, y) dx \\
    &\qquad \to
        \DG^*(fg)(y) - f(y) \DG^* g(y) . 
\end{align*}
Applying this to~\eqref{eqn:lineage_generator_Edt} with $g = \varphi_{T-s} \gamma$,
this suggests that the generator of the limiting process is
\begin{align} \label{eqn:heuristic_lineage_generator}
    \Lgen_s f
    &=
    \frac{r}{\varphi_{T-s}}
    \left\{
        \DG^*(\gamma \varphi_{T-s} f) - f \DG^*(\gamma \varphi_{T-s})
    \right\} .
\end{align}
This agrees with Theorem~\ref{thm:lineages}.

@@

The gneral set-up is as follows.
Each individual will be labeled with a ``level'',
which will be a number in $[0, \lambda]$.
When taking the limit, we will set $\lambda = N$.
\comment{Should we just use $N$ instead of $\lambda$? Or vice-versa? Change to $N$ but comment that when comparing ot other papers set $\lambda=N$.}
We will still encode the process embellished by these levels
as a point measure:
if the $i^\mathrm{th}$ individual's spatial location is $x_i$
and level is $u_i$, then we will write:
$$
    \lp = \sum_i \delta_{x_i, u_i} ,
$$
which is a measure on $\IR^d \times [0, \lambda]$.
Note that the each individual contributes mass 1 to the measure,
not $1/N$ as above.
In some sense, these levels do not affect the process:
the pushforward of this measure-valued ``spatial-level process'',
divided by $N$,
will agree with the purely spatial process defined above.
Furthermore, the levels of the individuals present at any spatial location
conditionally uniform on $[0, \lambda]$.
However, an individual's level in some sense encodes their future reproductive output:
individuals with lower levels tend to live longer, and have more offspring.
For more explanation of the set-up and how this is possible,
see \citet{etheridge/kurtz:2018} and \citet{kurtz/rodrigues:2011}.
Furthermore, the limiting spatial-level process will still be a point measure,
and so we explicitly retain the notion of individuals and lineages through to the 
large-population-density limit.

@@

We can now express our population processes using these levels:
\begin{align*}
    \eta_t = \frac{1}{\lambda} \sum_{a : \tau_a > t \;\&\; U_a(t) < \lambda} \delta_{X_a(t)}
    \qquad \text{and} \qquad
    \lp = \sum_{a : \tau_a > t \;\&\; U_a(t) < \lambda} \delta_{(X_a(t), U_a(t))} .
\end{align*}

@@

Now, let $\beta(x, \eta) = \meanq + 2 \Sigma \grad \log r(x, \eta)$,
and associate with each label an independent $d$-dimensional Brownian motion $W_a$,
and let $\covq = K K^T$.
Then, we can summarize the results above by saying that
in the infinite density limit,
the spatial locations and level of each line of descent $a$
behave for $\tau_a \le t < \sigma_a$ as
\begin{align*}
X_a(t)
    &=
    X_a(\tau_a)
    + \int_{\tau_a}^{t}
    \bigg(
        \beta(x, \eta(s)) ds
        +
        K dW_a(s)
    \bigg)
    ds  \\
U_a(t)
    &=
    U_a(\tau_a)
    + \int_{\tau_a}^{t}
    \bigg(
        \alpha \gamma(X_a(s),\eta(s))
        r(x,\eta(s)) U_a(s)^2
\\ &\qquad \qquad \qquad {}   
        -
        \left\{
            \gamma(X_a(s),\eta(s)) \DG r(X_a(s),\eta(s))
            + F(X_a(s), \eta(s))
        \right\}
        U_a(s)
    \bigg)
    ds .
\end{align*}
To complete the description,
for each label $a \in \labelspace$ let $R_a$ denote an independent Poisson process on $[0, \infty)^2$
with Lebesgue mean measure,
and as before order the points of each in some way.
If the population state at time 0 is $\eta(0)$,
then let $(x_i, u_i)$ be the points of a Poisson process
with mean measure on $\IR^d \times [0, \infty)$ a product of $\eta(0)$ and Lebesgue measure.
\comment{Here I've switched $t$ to be the absolute birth time
rather than relative to the birth time of the parent; I think it's clearer?
It, should be made to agree with the previous section.}
For each point begin a line of descent
with label $i$, location $X_i(0) = x_i$, level $U_a(0) = u_i$, and birth time $\tau_i = 0$.
Each point in the $R_a$ denotes a potential birth time for $a$:
if the $j^\text{th}$ point in $R_a$ is $(t, v)$,
$\tau_a \ge t$, and and $U_a(t) < \infty$,
then a new line of descent with label $a \concat j$ is produced,
with birth time $\tau_{a \concat j} = t$, location $X_{a \concat j}(t) = X_a(t)$, and level
$$
    U_{a \concat j}(t) = U_a(t)
    + \frac{v}{ 2 \alpha \gamma(X_a(t), \eta(t)) r(X_a(t), \eta(t)) } .
$$
\comment{Perhaps the conclusion of the Propostion would be something like:}
We can now express our population processes in the limit using these levels:
\begin{align*}
    \eta_t = \lim_{\lambda \to \infty} \frac{1}{\lambda} \sum_{a : \tau_a > t \;\&\; U_a(t) < \lambda} \delta_{X_a(t)}
    \qquad \text{and} \qquad
    \lp = \sum_{a : \tau_a > t} \delta_{(X_a(t), U_a(t))} .
\end{align*}
